The Impact of Alternation

### 2017-09-12 rebuttal.txt

We are grateful to all reviewers for their great work. The main points
of criticism are addressed as follows:

1. Why are our semi-algorithms performing slower than Includer [11] ?

First of all, Includer is specifically designed to solve language
inclusion problems between nondeterministic data automata, which are a
(significant) special case of alternation, whereas the semi-algorithms
presented in this paper can deal with the entire class of alternating
data automata, not just the ones resulting from inclusion problems. It
is common that procedures for more general problems are slower than
the ones for specific problems, although ideally, it is desirable to
have matching speed on the same class of problems.

As reviewer 2 points out, the generalized subset construction used by
Includer does not cause a huge blowup for the examples considered in
this paper and in [11]. Moreover, Includer uses the LIA interpolation
procedure of MathSAT5, which seems to be faster than the more generic
UFLIA interpolation of Z3 used here (UFLIA is needed in order to deal
with the boolean variables representing the states). In some cases, Z3
answers unsat almost immediately and then takes up to 5 seconds to
extract the interpolant. We believe that further experimenting with
different Z3 tactics and tuning the encoding of the interpolation
procedure, as well as considering a larger set of benchmarks than the
preliminary benchmarks used to test our first prototype will greatly
improve the performance of our tool.

2. What is the difference between the semi-algorithms in this paper
and the original ones of Blast [8] and Impact [15] ?

Both Blast and Impact are designed for checking emptiness (safety) of
nondeterministic data automata (programs), whereas our methods address
the emptiness problem for alternating data automata. A first
difference between our methods and [8,15] is a more sophisticated
Abstract Reachability Tree (ART) data structure (page 5, 2nd paragraph
of section 5), which requires labeling both nodes and edges with
formulae. However, on the positive side, our encoding of path formulae
(interpolation problems) allows the basic principles of [8,15] to be
carried over to the alternating setting (we represent path formulae as
conjunctions of implications and abstract sets of configurations as
combinations of boolean and data theory atoms), with some specific
differences listed below:

2.a) Difference with Blast [8]. Algorithm 1 (page 6) uses antichains
to prune the search space (lines 29-35). Using antichains is a
specific technique for classical finite-alphabet alternating
automata. We adapt this idea from [5] to optimize the predicate
abstraction emptiness checking of ADA.

2.b) Difference with Impact [15]. The original description of Impact
[15] is based on a nondeterministic application of the Cover, Refine
and Expand procedures, whereas Algorithm 2 (page 8) is a detailed
deterministic procedure that closely mirrors our implementation. In
particular, we chose to cover nodes lazily, only if their labels are
strenghtened by refinement (line 18).

3. Using the Lyndon interpolation theorem instead of post-images to
prove Proposition 1. This is a very interesting remark that we did not
consider initially and which allows to use positive interpolants
directly instead of post-images (quantifiers can be eliminated if the
data theory is LIA, but quantifier elimination has a non-negligeable
cost). If accepted, we are willing to change this point in the final
version, provided that we can find a published reference for the fact
that McMillan's Z3 interpolation procedure fits with Lyndon's theorem
(their CAV13, FMCAD11 or TCS05 papers on interpolation do not seem to
refer to it).

4. Several reviewers expressed concerns relative to the paper being
too theoretical and out of the scope of FMCAD. As authors of a paper
on formal verification techniques, we felt the need to define and
prove each point formally. However, the paper does not address issues
such as decidability or computational complexity of subclasses of ADA,
which could fit a more theoretical conference. Although interpolation
can be used as an argument for decidability (when suitable
restrictions are imposed, such as for symbolic alternating automata)
it is perhaps not a very good framework to derive complexity
bounds. In general, theoretical complexity proofs require different
techniques (e.g. reductions to other decidable problems whose
complexity is known) which are clearly out of the scope of the present
paper.

### 2017-10-19 tacas/draft.tex
ok

% \documentclass{llncs}
\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}

@@

\author{Radu Iosif and Xiao Xu \\
CNRS, Verimag, Universit\'e de Grenoble Alpes \\
Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\centerline{\includegraphics[scale=0.8]{Example.pdf}}
\vspace*{-\baselineskip}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

\paragraph{\em Example}
Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
$F=\set{q_3,q_4}$ and transitions:
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
\end{array}\]
The missing rules, such as $\Delta(q_0,b)$, are assumed to be
$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) using only atoms $x \geq k$
and $x = \overline{x} + k$, for $k \in \zed$, with the classical
interpretation of the function symbols on integers. Since reachability
of a control state is undecidable for AVASS \cite{LINCOLN92}, the
emptiness problem is undecidable for ADA.

@@

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
effect of reading the event $a$.

We generalize the post-image function to finite sequences of input
events, as follows:
\[\begin{array}{l}
\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
\end{array}\]
Then the emptiness problem for $\A$ becomes: does there exist $u \in
\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
Observe that, since we ask a satisfiability query, the final states of
$\A$ need not be constrained\footnote{ Since each state occurs
  positively in $\Accept{\A}(u)$, this formula has a model iff it has
  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

@@

In general, given an unsatisfiable conjunction $\Phi \equiv
\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
  \in X}$ of any sort, a solution of the \emph{interpolation problem}
$\Phi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
\item $I_0 \equiv \top$,  
%
\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
  I_i[X_i/X]$, for all $i \in [1,n]$, and
%
\item $I_n \equiv \bot$. 
\end{inparaenum}
In the following, we shall assume the existence of an interpolating
decision procedure for $\theory(\sig,\I)$.

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states
replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.

Moreover, we require that an interpolant
$\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
$\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
\in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
can always be built, as showed below:

\begin{proposition}\label{prop:positive-interpol}
  If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
  interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
  a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
  [0,n]$.
\end{proposition}

We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, these formulae lose track of the control history and are not
suitable for producing interpolants that relate states and data.

@@

We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.} \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose nodes are labeled with
formulae over-approximating the concrete sets of configurations, and a
covering relation between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample is eliminated by computing an interpolant and
adding its formulae to the set of predicates (cf. Lemma \ref{lemma:progress}).

@@

Formally, an ART is tuple $\Art =
\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
\item $N$ is a set of nodes, 
%
\item $E \subseteq N \times \Sigma \times
N$ is a set of edges, 
%
\item $\rootNode \in N$ is the root of the directed
tree $(N,E)$, 
%
\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
%
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
  with time-stamped formulae, and
%
\item $\lhd \subseteq N \times N$ is a set of
\emph{covering edges}. 
\end{compactitem}
%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
%% The replacement sets and time-stamped edge labels are defined below,
%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
%%   consider that the root node has zero depth.} $k$ in the tree
%% $(N,E)$ then we have:
%% \[\begin{array}{rcl}
%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
%% \end{array}\]

@@

With these considerations, Algorithm \ref{alg:predabs} uses a worklist
iteration to build an ART. We keep newly expanded nodes of $\Art$ in a
queue $\worklist$, thus implementing a breadth-first exploration
strategy, which guarantees that the shortest counterexamples are
explored first. When the search encounters a counterexample candidate
$u$, it is checked for spuriousness. If the counterexample is
feasible, the procedure returns a data word $w \in L(\A)$, which
interleaves the input events of $u$ with the data valuations from the
model of $\Accept{\A}(u)$ (since $u$ is feasible, clearly
$\Accept{\A}(u)$ is satisfiable). Otherwise, if $u$ is spurious, we
compute its pivot $p$ (line \ref{ln:pivot}), add the interpolants for
the least unfeasible suffix of $u$ to the set of predicates $\preds$,
remove and recompute the subtree of $\Art$ rooted at $p$.

@@

\begin{figure}[t!]
\begin{tabular}{cc}
\includegraphics[scale=0.5]{PA1.pdf} & \includegraphics[scale=0.5]{PA2.pdf} \\[-2mm]
\tiny{(a)} & \tiny{(b)} \\
\\
\includegraphics[scale=0.5]{PA3.pdf} & \includegraphics[scale=0.5]{PA4.pdf} \\[-2mm]
\tiny{(c)} & \tiny{(d)}
\end{tabular}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
\label{fig:pa}
\end{figure}

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{tabular}{c}
\includegraphics[scale=0.65]{IM1.pdf} \\
\tiny{(a)} \\
\includegraphics[scale=0.65]{IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.65]{IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\vspace*{-0.5\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact} (1/2)}
\label{fig:im1}
\vspace*{-\baselineskip}
\end{figure}

\begin{figure}[htb]
\centerline{\includegraphics[scale=0.65]{IM4.pdf}}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact} (2/2)}
\label{fig:im2}
\end{figure}

\paragraph{\em Example}
We show the execution of Algorithm \ref{alg:impact} on the automaton
from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im1}
(a). Since this node is uncovered, we check the spuriousness of the
counterexample $a$ and refine the label of the node to $q_1$. Since
the node is still uncovered, two successors, labeled with $\top$ are
computed, corresponding to the sequences $aa$ and $ab$, in Figure
\ref{fig:im1} (b). The spuriousness check for $aa$ yields the
interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
spurious, which changes the label of its endpoint from $\top$ to
$\bot$, and also covers it (depicted with a dashed edge). Since the
endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
Figure \ref{fig:im1} (c). Both sequences $aaa$ and $aab$ are found to
be spurious, and the enpoint of $aab$, whose label has changed from
$\top$ to $\bot$, is now covered. In the process, the label of $aa$
has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
the strenghtening with the interpolant from $aab$.  Finally, the only
uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
spurious, in Figure \ref{fig:im2}. The refinement of $aaab$ causes
the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
q_2$ and this node is now covered by $aa$. Since its successors are
also covered, there are no uncovered nodes and the procedure returns
$\true$. \qed

@@

\subsection{Proof of Lemma \ref{lemma:closure}}

@@

\subsection{Proof of Lemma \ref{lemma:safety-invariant}}

@@

\subsection{Proof of Lemma \ref{lemma:abstract-invariant}}

@@

\subsection{Proof of Proposition \ref{prop:positive-interpol}}

>>>>

\documentclass[10pt]{llncs}

% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}

@@

\author{Radu Iosif \and Xiao Xu}
\institute{CNRS, Verimag, Universit\'e de Grenoble Alpes \\
\email{\{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
}

@@

\begin{figure}[htb]
\begin{center}
\begin{minipage}{6cm}
\centerline{\includegraphics[scale=0.8]{Example.pdf}}
\centerline{(a)}
\end{minipage}
\begin{minipage}{6cm}
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y} \\\\
\end{array}\]
\centerline{(b)}
\end{minipage}
\end{center}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

\paragraph{\em Example}
Figure \ref{fig:ex} (a) depicts an ADA with input alphabet $\Sigma =
\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
$F=\set{q_3,q_4}$ and transitions given in Figure \ref{fig:ex} (b),
where missing rules, such as $\Delta(q_0,b)$, are assumed to be
$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) using only atoms $x \geq k$
and $x = \overline{x} + k$, for $k \in \zed$, with the classical
interpretation of the function symbols on integers. Since reachability
of a control state is undecidable for AVASS \cite{LINCOLN92}, 
emptiness is undecidable for ADA.

@@

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
effect of reading the event $a$. We generalize the post-image function
to finite sequences of input events, as follows:
\[\begin{array}{l}
\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
\end{array}\]
Then the emptiness problem for $\A$ becomes: does there exist $u \in
\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
Observe that, since we ask a satisfiability query, the final states of
$\A$ need not be constrained\footnote{ Since each state occurs
  positively in $\Accept{\A}(u)$, this formula has a model iff it has
  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

@@

Given an unsatisfiable conjunction of formulae $\psi_1 \wedge \ldots
\wedge \psi_n$, an \emph{interpolant} is a tuple of formulae
$\tuple{I_1,\ldots,I_{n-1},I_n}$ such that $I_n \equiv \bot$, $I_i
\wedge \psi_i \models_\T I_{i+1}$ and $I_i$ contains only variables
and function symbols that are common to $\psi_i$ and $\psi_{i+1}$, for
all $i \in [n-1]$. Moreover, by Lyndon's Interpolation Theorem
\cite{Lyndon59}, we can assume without loss of generality that every
boolean variable with at least one positive (negative) occurrence in
$I_i$ has at least one positive (negative) occurrence in both $\psi_i$
and $\psi_{i+1}$. In the following, we shall assume the existence of
an interpolating decision procedure for $\theory(\sig,\I)$ that meets
the requirements of Lyndon's Interpolation Theorem.

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states
replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.
%
%% Moreover, we require that an interpolant
%% $\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
%% $\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
%% \in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
%% can always be built, as showed below:
%
%% \begin{proposition}\label{prop:positive-interpol}
%%   If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
%%   interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
%%   a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
%%   [0,n]$.
%% \end{proposition}
%
We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, these formulae lose track of the control history and are not
suitable for producing interpolants that relate states and data.

@@

We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.} \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose nodes are labeled with
formulae over-approximating the concrete sets of configurations, and a
covering relation between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample is eliminated by computing an interpolant and
adding its formulae to the set of predicates (cf. Lemma
\ref{lemma:progress}). Formally, an ART is tuple $\Art =
\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
\item $N$ is a set of nodes, 
%
\item $E \subseteq N \times \Sigma \times
N$ is a set of edges, 
%
\item $\rootNode \in N$ is the root of the directed
tree $(N,E)$, 
%
\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
%
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
  with time-stamped formulae, and
%
\item $\lhd \subseteq N \times N$ is a set of
\emph{covering edges}. 
\end{compactitem}

@@

%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
%% The replacement sets and time-stamped edge labels are defined below,
%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
%%   consider that the root node has zero depth.} $k$ in the tree
%% $(N,E)$ then we have:
%% \[\begin{array}{rcl}
%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
%% \end{array}\]

@@

With these considerations, Algorithm \ref{alg:predabs} uses a worklist
iteration to build an ART. We keep newly expanded nodes of $\Art$ in a
queue $\worklist$, thus implementing a breadth-first exploration
strategy, which guarantees that the shortest counterexamples are
explored first. When the search encounters a counterexample candidate
$u$, it is checked for spuriousness. If the counterexample is
feasible, the procedure returns a data word $w \in L(\A)$, which
interleaves the input events of $u$ with the data valuations from the
model of $\Accept{\A}(u)$ (since $u$ is feasible, clearly
$\Accept{\A}(u)$ is satisfiable). Otherwise, $u$ is spurious and we
compute its pivot $p$ (line \ref{ln:pivot}), add the interpolants for
the least unfeasible suffix of $u$ to $\preds$, remove and recompute
the subtree of $\Art$ rooted at $p$.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{cc}
\includegraphics[scale=0.5]{PA1.pdf} & \includegraphics[scale=0.5]{PA2.pdf} \\[-2mm]
\tiny{(a)} & \tiny{(b)} \\
\\
\includegraphics[scale=0.5]{PA3.pdf} & \includegraphics[scale=0.5]{PA4.pdf} \\[-2mm]
\tiny{(c)} & \tiny{(d)}
\end{tabular}
\end{center}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
\label{fig:pa}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\begin{minipage}{6cm}
\begin{tabular}{c}
\includegraphics[scale=0.60]{IM1.pdf} \\
\tiny{(a)} \\
\includegraphics[scale=0.60]{IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.60]{IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\end{minipage}
\begin{minipage}{6cm}
\includegraphics[scale=0.60]{IM4.pdf}
\centerline{\tiny{(d)}}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact}}
\label{fig:im}
\end{figure}

\paragraph{\em Example}
We show the execution of Algorithm \ref{alg:impact} on the automaton
from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im}
(a). Since this node is uncovered, we check the spuriousness of the
counterexample $a$ and refine the label of the node to $q_1$. Since
the node is still uncovered, two successors, labeled with $\top$ are
computed, corresponding to the sequences $aa$ and $ab$, in Figure
\ref{fig:im} (b). The spuriousness check for $aa$ yields the
interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
spurious, which changes the label of its endpoint from $\top$ to
$\bot$, and also covers it (depicted with a dashed edge). Since the
endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
Figure \ref{fig:im} (c). Both sequences $aaa$ and $aab$ are found to
be spurious, and the enpoint of $aab$, whose label has changed from
$\top$ to $\bot$, is now covered. In the process, the label of $aa$
has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
the strenghtening with the interpolant from $aab$.  Finally, the only
uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
spurious, in Figure \ref{fig:im} (d). The refinement of $aaab$ causes
the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
q_2$ and this node is now covered by $aa$. Since its successors are
also covered, there are no uncovered nodes and the procedure returns
$\true$. \qed

@@

\subsection*{Proof of Lemma \ref{lemma:closure}}

@@

\subsection*{Proof of Lemma \ref{lemma:safety-invariant}}

@@

\subsection*{Proof of Lemma \ref{lemma:abstract-invariant}}

@@

%% \subsection*{Proof of Proposition \ref{prop:positive-interpol}}
